### 2022-04-21 misc/expo/tensorrank.tex
Writing stuff

\def\FF{{\bf F}}
\def\TT{{\bf T}}
\def\bar{\overline}
\def\hat{\widehat}
\def\norm#1{|\!|#1|\!|}
\def\bignorm#1{\big|\!\big|#1\big|\!\big|}
\def\Norm#1{\Big|\!\Big|#1\Big|\!\Big|}
\def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}

@@

\maketitle{The analytic rank of a tensor}{by}{Marcel K. Goh}{10 April 2022}

>>>>

\def\FF{{\bf F}}
\def\T{{\rm T}}
\def\bias{\op{\rm bias}}
\def\prk{\op{\rm prk}}
\def\ark{\op{\rm ark}}
\def\bar{\overline}
\def\hat{\widehat}
\def\norm#1{|\!|#1|\!|}
\def\bignorm#1{\big|\!\big|#1\big|\!\big|}
\def\Norm#1{\Big|\!\Big|#1\Big|\!\Big|}
\def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}
\def\x{{\bf x}}
\def\y{{\bf y}}

@@

\maketitle{The analytic rank of a tensor}{by}{Marcel K. Goh}{13 April 2022}

\floattext4.5 \ninebf Note. \ninepoint These notes are more or less a retelling of
some results in a 2019 paper entitled ``The analytic rank of tensors and its applications'' by
S.~Lovett. I go more deeply into definitions (for my own sake) so for other students this may be easier to
follow than the original paper.

\bigskip

\advsect Introduction

\noindent
There are various compatible definitions of the {\it rank} of a matrix. The one that
extends most easily to the context of tensors, which we define later, is the following.
An $m\times n$ matrix $A$ is said to be {\it rank one} if there exist vectors $u\in \FF^m$
and $v\in \FF^n$ such that $A = uv^\T$. The {\it rank} of a general matrix $A$ is the minimum
number $k$ such that we can write $A = A_1 + \cdots + A_k$, where $A_i$ is a rank one matrix
for all $1\le i\le k$.

In a first course on linear algebra, one usually learns that the rank of a matrix is the
dimension of its column (or row) space. To see that the above definition of rank is equivalent,
let $A$ be a rank-$k$ matrix and let $B = \{b_1, \ldots, b_k\}$
be a basis of its column space. Since every column of $A$
can be written as a linear combination of vectors in $B$, so there is a $k\times n$ matrix $C$
such that $A = BC$. Now letting $c_1, \ldots, c_k$ be the rows of $C$, we have
$$A = b_1{c_1}^\T + \cdots + b_k{c_k}^\T,$$
so the rank of $A$ is at most $k$. On the other hand, if
$$A = u_1{v_1}^\T + \cdots + u_{k'}{v_{k'}}^\T,$$
for some $k'< k$, then for all $x \in \FF^n$, we have
$$Ax = u_1{v_1}^\T x + \cdots + u_{k'}{v_{k'}}^\T x.$$
Since ${v_i}^\T x$ is a scalar for all $1\le i\le k'$, we conclude that $u_1, \ldots, u_{k'}$
span the image of $A$ and the column space of $A$ is at most $k'<k$.

Any $m\times n$ matrix $A$ gives a bilinear map from $\FF^m \times \FF^n$ to $\FF$ by taking
$(x,y)\mapsto xAy$. We extend this to more than two vector spaces by defining an {\it order-$d$ tensor}
to be a multilinear map $T : V_1\times\cdots\times V_d\to \FF$, where $V_i$ is a vector space over $\FF$
for all $1\le i\le d$. From here on out,
we restrict ourselves to the case where each $V_i$ has the same dimension $n$ and can thus be identified
with $\FF^n$. Then there exist $n^d$ scalars $\{T_{j_1,\ldots,j_d}\}_{j_1,\ldots,j_d\in [n]}$
such that for all $x_1,\ldots,x_d\in \FF^n$,
$$ T(x_1, \ldots, x_d) = \sum_{j_1,\ldots,j_d\in [n]} T_{j_1,\ldots,j_d} x_{1,j_1}\cdots x_{d,j_d},$$
where $x_{i, k}$ denotes the $k$th component of the vector $x_i$. There is thus a one-to-one correspondence
between order-$d$ tensors and $d$-dimensional arrays of scalars (in our setting each dimension has size
$n$).
If $T$ is an order-$d$ tensor and $T'$ is an order-$d'$ tensor, then we can form a tensor of order $d+d'$
from the $(d+d')$-dimensional array of scalars
$$\{T_{i_1,\ldots,i_d}T'_{j_1,\ldots,j_{d'}}\}_{i_1,\ldots,i_{d},j_1,\ldots,j_{d'}\in [n]}.$$
This tensor is denoted $T\otimes T'$ and is called the {\it tensor product} of $T$ and $T'$.

Now we say that an order-$d$ tensor $T$ is {\it partition rank one} if there exists $A\subseteq [d]$
with $0 < |A| < d$, as well as an order-$|A|$ tensor $T_1$ and an order-$(d-|A|)$ tensor $T_2$ such
that $T$ can be written as
$$ T(x_1, \ldots, x_d) = T_1(x_i : i\in A)T_2(x_i : i\notin A).$$
The {\it partition rank} $\prk(T)$ of a
general tensor $T$ is the minimum $k$ such that $T$ can be written as a sum
of $k$ partition rank one tensors. Note that in the case $d=2$ this reduces to the ordinary matrix rank.

\medskip\boldlabel The cap set problem. The partition rank was introduced to study the cap-set problem,
and here we shall sketch how it applies. A {\it cap set} is a subset $A\subseteq \FF_3^n$ such that for
every triple $(x,y,z)\in A$ of pairwise distinct elements, $x+y+z\ne 0$. In 1984, T.~C.~Brown
and J.~P.~Buhler showed that, loosely speaking, cap sets have zero density.

\parenproclaim Theorem C (Brown--Buhler{\rm, 1986}). For every $\delta>0$
there exists $n$ such that every subset $A\subseteq \FF_3^n$ with $|A|\ge \delta 3^n$ contains three pairwise
distinct elements $x$, $y$, and $z$ with $x+y+z=0$.

A later paper by R.~Meshulam gave better quantitative
bounds on $n$ with respect to $\delta$; namely, it was shown that we need only take $n > 2/\delta$.
This means that if $A$ is a cap set in $\FF_3^n$, then $|A| \le 2\cdot 3^n/n$. However, it was long
suspected that this bound could be improved to $|A| \le O(c^n)$ for some $c<3$. This was finally proved
in 2017 by J.~S.~Ellenberg and D.~Gijswijt, and T.~Tao showed in a blog post (dated 18 May 2016)
that thir proof can be restated in terms of the partition rank of a function in $3$ variables. This
can actually be modified so the function is a $3$-tensor, but to just get the general idea, let
us extend our definition of partition rank to general functions of three variables temporarily.

Given $A\subseteq \FF_3^n$, let $T : V^3 \to \FF_3$, where $V = {\FF_3}^{\FF_3^n}$, be given by
$$f(e_a,e_b,e_c = \cases{1, & if $a+b+c=0$;\cr 0, & otherwise,}$$
for basis vectors $e_v$ and extended to all other vectors by linearity. (The function $e_v$ has
$e_v(v) = 1$ and $e_v(x)=0$ for all $x\ne v$.)
Now for a tensor $T : (\FF^X)^d \to \FF$, we say that a set $A\subseteq X$ is an {\it independent
set in $T$} if for all $i_1,\ldots,i_d\in A$, the condition that the coefficient $T_{i_1,\ldots,i_d}$
be nonzero is equivalent to $i_1=\cdots=i_d$.
We then give an upper bound on the size of a cap set by proving that
\medskip
\item{i)} if $A$ contains no nontrivial solutions to $x+y+z=0$, then $A$ is an independent set in $T$;
\smallskip
\item{ii)} if $A$ is an independent set in $T$ then $\prk(T)\ge |A|$; and
\smallskip
\item{iii)} the partition rank of $T$ is low.
\medskip
In these notes, we aim to show that this general strategy may be performed with the partition rank
replaced by something called the analytic rank.

\medskip\boldlabel The analytic rank.
In a 2011 paper, W.~T.~Gowers and J.~Wolf introduced another definition of rank that is Fourier-analytic
in nature. Now we require the field $\FF$ to be finite, and let $\chi:\FF\to\CC$ be any nontrivial
additive character. Recall that for such a character, $\ex_{a\in \FF} \chi(a) = 0$.
The {\it bias} of a tensor $T : V^d\to \FF$ is the average
$$\bias(T) = \ex_{x\in V^d} \chi\bigl(T(x)\bigr).$$
Note that if $T$ is a linear form (i.e., an order-$1$ tensor) that is not identically zero,
then $\bias(T) = 0$, since we can bring the sum inside all three functions and the sum over all elements
a vector space over a finite field is zero. If $T$ is identically zero then $\bias(T) = 1$.
Now to see that the bias of a tensor is always in $(0,1]$, note that if we fix any $(x_2, \ldots, x_d)\in V^{d-1}$,
then $T(x_1,x_2,\ldots,x_d)$ becomes a linear form (order-$1$ tensor) in $x_1$ and
$$\eqalign{
\bias(T)
&= \ex_{x_2,\ldots,x_d\in V} \ex_{x_1\in V} \chi\bigl(T(x_1,\ldots,x_d)\bigr)\cr
&=\pr_{x_2,\ldots,x_d\in V} \bigl\{ T(x_1,\ldots,x_d) \equiv 0\bigr\},\cr
}$$
from our earlier observation about order-$1$ tensors.

The {\it analytic rank} is defined to be the quantity
$$\ark(T) = -\log_{|\FF|} \bias(T);$$
since $\bias(T) \in (0,1]$ we have $\ark(T)\ge 0$. In the case of order-$2$ tensors, the analytic rank
is once again equivalent to ordinary matrix rank. To see this, suppose that $T : (\FF^n)^2\to \FF$
is defined as $T(x,y) = \sum_{i=1}^r x_iy_i$. Then $\bias(T)$ is the probability that, fixing
$y$, the linear form $T(x,y)$ is identically zero. This is equivalent to every coordinate of $y$ being
zero, which happens with probability $1/|\FF|^r$, and hence we see that $\ark(T) = r$.

\advsect Subadditivity of analytic rank

The goal of this section is to prove that if $T$ and $S$ are tensors, then $\ark(T+S)\le \ark(T)+\ark(S)$.
Our first small lemma is the following.

\proclaim Lemma \advthm. Let $W_0, W_1,\ldots,W_n : \FF^m\to \FF$ be functions. Let
functions $A,B : \FF^n\times \FF^m\to \FF$ be given by
$$A(x,y) = \sum_{i=1}^n x_i W_i(y)\qquad\hbox{and}\qquad B(x,y) = A(x,y) + W_0(y).$$
Then
$$\bigl| \bias(B)\bigr| \le \bias(A).$$

\proof We expand
$$\bias(B) = \ex_{x\in \FF^n, y\in \FF_m} \chi\bigl(B(x,y)\bigr)
= \ex_{y\in \FF^m} \one_{W_1(y) = \cdots = W_n(y) = 0} \cdot \chi\bigl(W_0(y)\bigr)$$
and by the triangle inequality,
$$\bigl|\bias(B)\bigr| \le \ex_y \one_{W_1(y) = \cdots = W_n(y) = 0} = \bias(A).\noskipslug$$

This lemma is used to prove a bound on the bias of a certain sum of tensors.
First, we introduce some notation. With some $d$ fixed, we let $\x = (x_1,\ldots,x_d)$ and
similarly for $\y = (y_1,\ldots,y_d)$. Then for
$I\subseteq [d]$, define $I^c = [d]\setminus I$ and let $\x_I = (x_i : i\in I)$.

\proclaim Lemma \advthm. Let $d\ge 1$ and for each $I\subseteq [d]$, let $R_I : V^I\to \FF$
be an order-$|I|$ tensor. Consider the function
$$R(\x) = \sum_{I\subseteq [d]} R_I(x_I).$$
Then
$$\bigl|\bias(R)\bigr| \le \bias(R_{[d]}).$$

\proof Fix some $i\in [d]$ and write $R(\x)$ as
$$R(\x) = \sum_{I \ni i} R_I(x_I) + \sum_{I\not\ni i} R_I(x_I).$$
Setting $x = x_i$ and $y = \x_{[d]\setminus\{i\}}$, the first sum has the form
$$\sum_{i=1}^n x_iW_i(y)$$ and the second sum does not depend on $x$ at all, so we can set it
to be $W_0(y)$. The previous lemma now tells us that
$$\bigl|\bias(R)\bigr| \le \bias\Bigl(\sum_{I\ni i} R_I(\x_I)\Bigr).$$
Now iterate this with $i=d$ all the way down to $i=1$ (replacing $d$ by $d-1$ each time)
to get the statement of the lemma.\slug

\proclaim Theorem \advthm. Let $T,S: V^d\to \FF$ be order-$d$ tensors. Then
$$\ark(T+S) \le \ark(T) + \ark(S).$$

\section References
